%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% Writing --- and Publishing -- Novels with LaTeX
% By Tammy Cravit, tammymakesthings@gmail.com
%
% Version     : 1.00
% Created at  : <Thu Sep 30 06:07:37 2021>
% Last updated: Time-stamp: <2021-10-14 06:29:37 tammy>
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% File Name   : chapters/01_introduction.tex
% Content     : Chapter 1: Introduction
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% Copyright (c) 2021, Tammy Cravit. All rights reserved.  This work is
% licensed under a Creative Commons Attribution-NonCommercial-ShareAlike 4.0
% International License.
%
% You are free to:
%
%     Share — copy and redistribute the material in any medium or format
%
%     Adapt — remix, transform, and build upon the material
%
%     The licensor cannot revoke these freedoms as long as you follow the
%     license terms.
%
% Under the following terms:
%
%     Attribution — You must give appropriate credit, provide a link to the
%     license, and indicate if changes were made. You may do so in any
%     reasonable manner, but not in any way that suggests the licensor
%     endorses you or your use.
%
%     NonCommercial — You may not use the material for commercial purposes.
%
%     ShareAlike — If you remix, transform, or build upon the material, you
%     must distribute your contributions under the same license as the
%     original.
%
%     No additional restrictions — You may not apply legal terms or
%     technological measures that legally restrict others from doing anything
%     the license permits.
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\chapter{Introduction} \label{chap:introduction}

In today's world, authors who plan to publish their own work are faced with a
daunting set of challenges. How do we maintain print, ebook, and other formats
of our novels? How do we format and typeset these works, and maintain all
those different versions and formats, without spending endless hours on the
process of design and formatting?

The answer --- at least for me --- turns out to lie in a 50-year-old piece of
software that can automate a lot of the layout and typesetting work. \TeX{} (and
its more user-friendly author oriented progeny, \LaTeX) were first developed
in the 1970s to typeset mathematical and scientific texts. Over the years,
they've grown to allow typesetting all manner of documents, including fiction
and poetry, and to produce PDF, ebook, and (with a bit of help) other output
formats such as Word and HTML. A vast ecosystem of add-on packages can help
you typeset nearly any kind of document, including sheet music.

The other nice thing about this solution is that you're not locked into any
one software vendor's proprietary tools. \LaTeX{} documents, just like the HTML
files that make up Web sites, are simply text files. You can edit them on any
platform, with whatever tools you like. You don't have to worry about your
work becoming inaccessible if a new version of your editing tool comes out or
a software product is discontinued. This is also a great way to
``future-proof'' your writing --- your documents don't depend on any
particular computer platform or software vendor still being around 10 or 20
years from now.

Oh, and did I mention it's all free?

This article will describe my process for writing, editing, and publishing
novels with \LaTeX. There's certainly more than one way to do things, so this
is a description of \emph{my} process, not the ``one true way'' to tackle this
problem. I'll provide a few resources at the end so you can learn more and see
how other authors are using the \LaTeX{} ecosystem to publish their works.

Ready? Let's get started!

%%% Local Variables:
%%% fill-column: 78
%%% TeX-master: "../latex_tooling_fiction_writing"
%%% End:
